\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,graphicx,xcolor,url,hyperref,breakurl,enumerate}

\begin{document}
\begin{center}\Large\bf
  Homework 3\\
  Data Science\\
  PCS – Uconn, 2022\\
  Due Date: July 14, 2022\\
  (by 11:59pm, via HuskyCT)
\end{center}

\begin{enumerate}[1.]
\item We have twenty-five students and three professors in our class. Suppose
  that each year has 365 days. Use simulation to find the probability of the
  following event. 
\begin{itemize}
\item \emph{Exactly two} out of twenty-eight people have a common birthday and
  other people have different birthdays.
\end{itemize}

\item A King offers a death row prisoner a chance to live by playing a simple
  game. He gives 50 black marbles, 50 white marbles, and 2 empty bowls to the
  prisoner. The King asks the prisoner to divide these 100 marbles into these 2
  bowls in any way as long as all the marbles are used. He will then blindfold
  the prisoner and mix the bowls around. After that, the prisoner can choose one
  bowl and pick one marble randomly.  If the marble is white the prisoner will
  live, but if the marble is black the prisoner will die. Use simulation to find
  the probability of living for the following ways of dividing.
  \begin{enumerate}[a.]
  \item Put the 50 white balls in one bowl and the 50 black balls in another
    bowl.
  \item Put one white ball in one bowl and the other 99 bowl balls in another
    bowl.
  \end{enumerate}

\end{enumerate}

\end{document}
